\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-09-07 paper.qmd
Updated PBMC3k description

To determine whether the UMAP representation with the specific (hyper-)parameter choice suggested by @chen2023 preserves the data structure present in $9\text{-}D$ (@fig-model-pbmc-author-proj (b)), we visualize the model constructed with UMAP overlaid on the $9\text{-}D$ data. The @fig-NLDR-variety (a) shows three well-separated clusters with big separations. However, as shown in @fig-model-pbmc-author-proj (a2), there is no big separation between three clusters in $9\text{-}D$. Therefore, the suggested UMAP representation (@fig-NLDR-variety (a)) does not accurately represent the structure of PBMC3k dataset.

@@

The MSE for different binwidths ($a_1$) using tSNE, UMAP, PHATE, PaCMAP, and TriMAP with different (hyper-)parameter settings (@fig-pbmc-mse) were calculated. It is found that tSNE (@fig-mnist-mse (a)) provide the most reasonable representation for the digit 1 dataset, showing universally best. However, \gD{} representation shows a big non-linear cluster and a small cluster with a small gap (@fig-tsne-best).

In order to find a reasonable NLDR representation for the PBMC3k dataset, the MSE for different $a_1$ using UMAP, tSNE, PHATE, PaCMAP, and TriMAP with different (hyper-)parameter settings (@fig-pbmc-mse) were calculated. tSNE with a perplexity value set to $30$, the default parameter setting, is considered as a reasonable representation for the PBMC3k dataset. As shown in @fig-NLDR-variety (e), there are three well-separated clusters, although the separation between the clusters are small. Additionally, non-linear structures can also be observed within the clusters (@fig-NLDR-variety (e)).
The visualization of the *model-in-the-data-space* provides evidence for this (@fig-pbmc2-sc). Therefore, tSNE with perplexity $30$ accurately captures the data structure of the PBMC3k dataset, which is better than UMAP.

>>>>

The @fig-NLDR-variety (a) shows three well-separated clusters with big separations. However, as shown in @fig-model-pbmc-author-proj (a2), there is no big separation between three clusters in $9\text{-}D$. Therefore, the suggested UMAP representation (@fig-NLDR-variety (a)) does not accurately represent the structure of PBMC3k dataset.

As shown in @fig-NLDR-variety, layout c, d, and e that show small separations between clusters are universally optimal. But layout d with little separations perform well at tiny binwidth and poorly as binwidth increases. Also, layout e performs well at large binwidths. The layout c is chosen as the best, since it used the UMAP with a different (hyper-) parameter settings.

The visualization of the model generated for the selected layout in $9\text{-}D$ shows edges between the clusters (@fig-mnist-tri-proj (b2)). This provides evidence for the small separations between clusters. Furthermore, the data does not uniformally distributed along the clusters and therefore shows densed points (@fig-mnist-tri-proj (b2). Also, the clusters are non-linear shaped (@fig-mnist-tri-proj (b3).

<!-- To determine whether the UMAP representation with the specific (hyper-)parameter choice suggested by @chen2023 preserves the data structure present in $9\text{-}D$ (@fig-model-pbmc-author-proj (b)), we visualize the model constructed with UMAP overlaid on the $9\text{-}D$ data.  -->

@@

<!-- The MSE for different binwidths ($a_1$) using tSNE, UMAP, PHATE, PaCMAP, and TriMAP with different (hyper-)parameter settings (@fig-pbmc-mse) were calculated.  -->

<!-- It is found that tSNE (@fig-mnist-mse (a)) provide the most reasonable representation for the digit 1 dataset, showing universally best. However, \gD{} representation shows a big non-linear cluster and a small cluster with a small gap (@fig-tsne-best).  -->

<!-- In order to find a reasonable NLDR representation for the PBMC3k dataset, the MSE for different $a_1$ using UMAP, tSNE, PHATE, PaCMAP, and TriMAP with different (hyper-)parameter settings (@fig-pbmc-mse) were calculated. tSNE with a perplexity value set to $30$, the default parameter setting, is considered as a reasonable representation for the PBMC3k dataset. As shown in @fig-NLDR-variety (e), there are three well-separated clusters, although the separation between the clusters are small. Additionally, non-linear structures can also be observed within the clusters (@fig-NLDR-variety (e)). -->
<!-- The visualization of the *model-in-the-data-space* provides evidence for this (@fig-pbmc2-sc). Therefore, tSNE with perplexity $30$ accurately captures the data structure of the PBMC3k dataset, which is better than UMAP. -->

### 2024-09-07 paper.tex
Updated PBMC3k description

To determine whether the UMAP representation with the specific
(hyper-)parameter choice suggested by \citet{chen2023} preserves the
data structure present in \(9\text{-}D\)
(Figure~\ref{fig-model-pbmc-author-proj} (b)), we visualize the model
constructed with UMAP overlaid on the \(9\text{-}D\) data. The
Figure~\ref{fig-NLDR-variety} (a) shows three well-separated clusters
with big separations. However, as shown in
Figure~\ref{fig-model-pbmc-author-proj} (a2), there is no big separation
between three clusters in \(9\text{-}D\). Therefore, the suggested UMAP
representation (Figure~\ref{fig-NLDR-variety} (a)) does not accurately
represent the structure of PBMC3k dataset.

@@

The MSE for different binwidths (\(a_1\)) using tSNE, UMAP, PHATE,
PaCMAP, and TriMAP with different (hyper-)parameter settings
(Figure~\ref{fig-pbmc-mse}) were calculated. It is found that tSNE
(Figure~\ref{fig-mnist-mse} (a)) provide the most reasonable
representation for the digit 1 dataset, showing universally best.
However, \gD{} representation shows a big non-linear cluster and a small
cluster with a small gap (Figure~\ref{fig-tsne-best}).

In order to find a reasonable NLDR representation for the PBMC3k
dataset, the MSE for different \(a_1\) using UMAP, tSNE, PHATE, PaCMAP,
and TriMAP with different (hyper-)parameter settings
(Figure~\ref{fig-pbmc-mse}) were calculated. tSNE with a perplexity
value set to \(30\), the default parameter setting, is considered as a
reasonable representation for the PBMC3k dataset. As shown in
Figure~\ref{fig-NLDR-variety} (e), there are three well-separated
clusters, although the separation between the clusters are small.
Additionally, non-linear structures can also be observed within the
clusters (Figure~\ref{fig-NLDR-variety} (e)). The visualization of the
\emph{model-in-the-data-space} provides evidence for this
(\textbf{?@fig-pbmc2-sc}). Therefore, tSNE with perplexity \(30\)
accurately captures the data structure of the PBMC3k dataset, which is
better than UMAP.

>>>>

The Figure~\ref{fig-NLDR-variety} (a) shows three well-separated
clusters with big separations. However, as shown in
Figure~\ref{fig-model-pbmc-author-proj} (a2), there is no big separation
between three clusters in \(9\text{-}D\). Therefore, the suggested UMAP
representation (Figure~\ref{fig-NLDR-variety} (a)) does not accurately
represent the structure of PBMC3k dataset.

As shown in Figure~\ref{fig-NLDR-variety}, layout c, d, and e that show
small separations between clusters are universally optimal. But layout d
with little separations perform well at tiny binwidth and poorly as
binwidth increases. Also, layout e performs well at large binwidths. The
layout c is chosen as the best, since it used the UMAP with a different
(hyper-) parameter settings.

The visualization of the model generated for the selected layout in
\(9\text{-}D\) shows edges between the clusters
(Figure~\ref{fig-mnist-tri-proj} (b2)). This provides evidence for the
small separations between clusters. Furthermore, the data does not
uniformally distributed along the clusters and therefore shows densed
points (Figure~\ref{fig-mnist-tri-proj} (b2). Also, the clusters are
non-linear shaped (Figure~\ref{fig-mnist-tri-proj} (b3).

@@
